---
title: "Class Work - Week 1"
author: "Farhad Shah"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    css: !expr here::here("./StylesTemplates/jr_19.css")
---

```{r setup, include=F}
knitr::opts_chunk$set(fig.width=8, fig.height=5, fig.align = 'left', fig.path='Figs/', cache.path='Cache/', eval=T, echo=T, tidy=TRUE,  cache=F, message=F, warning=F, stringsAsFactors=F, yaml.eval.expr = TRUE)  

library(xfun)
pkg_attach("tidyverse", "here", "kableExtra", "RCurl", "psych", "knitr", "RColorBrewer", 'dplyr', install=T)

jamie.theme <- theme_bw() + theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank(), panel.background = element_blank(), legend.title= element_blank())  #custom theme ggplot2


print.me <- function(x, ...) {
if (nrow(x) > 200){
   len <- 200 
    } else {
       len <- (nrow(x))
}
   x[1:len,] %>%
   kbl(digits=2, align= 'l', booktabs=T) %>%
   kable_styling(fixed_thead = T) %>%
   kable_paper("striped", full_width = T, html_font = "Helvetica", font_size = 12) %>%
   row_spec(0, color = "white", background = "#5b705f", font_size = 12) %>%
   scroll_box(width = "700px", height = "500px") %>%
   asis_output()
}

registerS3method("knit_print", "data.frame", print.me)
#to_R <- read.csv(here("data", "MyRaw.txt")) ---- using 'here' to read in data
```

# My First Homework
I'm going to produce a string called marry had a little lamb in one code block. Then I'm going to split the string to one word per row. Then I'm going to add a variable called Nletters that counts the letters in each word. The other part of my homework is to style my css to my personal needs.

## It's getting ugly already
```{r}
#print a sequence of numbers 1:10
myseq <- seq(1:10)
print(myseq)
```
```{r}
#produce a string
sentence <- "mary had a little lamb"

#split the string
words <- unlist(strsplit(sentence, " "))

# add a variable to count letters in each word
df <- data.frame(
  word = words,
  stringsAsFactors = FALSE
)
df <- df %>%
  mutate(Nletters = nchar(word))
print(df)
```
# Next challenge
Now I'm gonna add a new string in which the sentence "marry had a little lamb" repeats 10000 times in rows

```{r}
# Create a string repeated 10,000 times (one per row)
df_sentences <- data.frame(
  sentence = rep("marry had a little lamb", 10000)
)

head(df_sentences)
```

